Influence of Time Integration Methodology on Double Detonation Type Ia Supernova Nucleosynthesis

### 2023-06-25 subch_tols/paper.tex
add plot
update paper

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We explore the coupling of hydrodynamics and reactions in simulations of
the double detonation model for Type Ia supernovae.  When assessing the
convergence of simulations, the focus is usually on spatial resolution.
However, the method of coupling the physics together as well as the
tolerances used in integrating a reaction network also play an important role.
In this paper, we explore how the choices made in integrating the reaction
portion of a simulation of the double detonation model of Type Ia supernovae
influences the nucleosynthesis.  We find that the stronger coupling
afforded by the simplified spectral deferred corrections method introduced
previously provides a more accurate and efficient result than standard
operator splitting.
\end{abstract}

@@

There are many different progenitor models for Type Ia supernova that
are currently being explored.  The double detonation model
\citep{fink:2007} involves a sub-Chandrasekhar mass white dwarf with a
thin accreted He layer.  A detonation in the He layer propagates
around the star while sending a compression wave converging toward the
center.  This compression wave ignites a carbon detonation that then
propagates outward from the center of the white dwarf.  Many groups
have modeled this scenario, varying the initial conditions,
resolution \citep{rivas:2022}, and reaction networks to understand the robustness of this
model as well as its connection to observations.

\citep{townsley:2019} \citep{gronow:2020}, \citep{gronow:2021}, \citep{glasner:2018}, \citep{roy:2022}, \citep{shenbildsten:2014},

These events are modeled using the Euler equations with reactive and
gravitational sources:
\begin{eqnarray*}
\ddt{\rho} + \nabla \cdot (\rho \Ub) &=& 0 \\
\ddt{(\rho \Ub)} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &=& \rho \gb \\
\ddt{(\rho E)} + \nabla \cdot (\rho \Ub E + \Ub p) &=& \rho \Ub \cdot \gb + \rho \dot{S} \\
\ddt{(\rho X_k)} + \nabla \cdot (\rho \Ub X_k) &=& \rho \omegadot_k
\end{eqnarray*}
where $\rho$ is the mass density, $\Ub$ is the velocity, $E$ is the specific
total energy, $p$ is the pressure, and $X_k$ are the mass fractions
of the nuclei that react.  The sources are the gravitational acceleration,
$\gb$, which is obtained by solving the the Poisson equation
\begin{equation}
\nabla^2 \Phi = 4 \pi G \rho
\end{equation}
(where $G$ is Newton's constant) for the gravitational potential, $\Phi$, and defining $\gb = -\nabla \Phi$.
The reaction network provides the species creation rate, $\omegadot$ and the energy generation rate, $\dot{S}$.
Finally, the system is closed via the equation of state:
\begin{equation}
p = p(\rho, e, X_k)
\end{equation}
where $e$ is the specific internal energy, obtained as $e = E - |\Ub|^2/2$.

Simulation codes use different methods to advance this system of equations.  For hydrodynamics,
explicit-in-time integration is usually used, with the timestep, $\Delta t$, restricted by the Courant limit:
\begin{equation}
\Delta t = C \min_{i,j,k} \left \{ \frac{\Delta x}{|\Ub| + c} \right \}
\end{equation}
where $\Delta x$ is the width of a grid zone and $c$ is the sound
speed, and the minimum is taken over all zones on the grid.  Here, the
Courant number, $C$ is required to be $C < 1$ for stability.
Reactions, on the other hand, tend to be stiff, and are usually evolved
using implicit-in-time integration, with many small steps taken to
make up the hydrodynamics timestep $\Delta t$.

@@

When using operator-splitting, the density remains constant, since there
is no reaction source in the mass continuity equation.  The mass
fractions, $X_k$, and energy evolve according to:
\begin{eqnarray}
\frac{dX_k}{dt} &= \omegadot_k \\
\frac{de}{dt} &= \dot{S} \\
\end{eqnarray}
where $\omegadot_k$ is the creation rate of species $k$ due to nuclear
reaction.  In the simplest approximation of operator splitting, the
evolution of the mass fractions is solved alone, without integrating
the temperature or energy.  This is the form that was used in FLASH
\cite{flash}.  A more accurate operator splitting also includes the
energy evolution (perhaps in terms of temperature).  This will require
calling the equation of state each time we evaluate the righthand side
of the ODE system, increasing the cost.  However, as we showed in
\citet{strang_rnaas}, this is needed to get second-order convergence.

Detonations can be difficult to model numerically.  Time step cuts \MarginPar{ref}.
For the double detonation scenario, the ignition of the second detonation
can be very sensitive to how the simulation is run \MarginPar{refs}.
Artifically limiting the rates \MarginPar{ref}

In \citet{castro_simple_sdc}, we introduced a method based on the
ideas of spectral deferred corrections for coupling reactions and
hydrodyamics (that we called ``simplified-SDC'').  Here, the overall
time-integration is done iteratively, with the hydrodynamics seeing an
explicit reactive source, and the reaction update evolving an ODE
system that includes a piecewise-constant-in-time advective source,
$\Adv{\Uc}^{n+1/2}$,
\begin{equation}
\ddt{\Uc} = \Rb(\Uc) + \Adv{\Uc}^{n+1/2}
\end{equation}

In addition to the time-integration strategy, integrating the reaction
ODE system requires specifying tolerances.  Unfortunately, the tolerances
used are not normally reported in papers, and in many cases, may not
be easily controlled by a code user at runtime.

@@

@@

\software{\amrex~\citep{amrex_joss},
          \castro~\citep{castro,castro_joss},
          GCC (\url{https://gcc.gnu.org/}),
          helmeos \citep{timmes_swesty:2000},
          linux (\url{https://www.kernel.org/}),
          matplotlib (\citealt{Hunter:2007}, \url{http://matplotlib.org/}),
          NumPy \citep{numpy,numpy2},
          pynucastro \citep{pynucastro,pynucastro2},
          python (\url{https://www.python.org/}),
          SymPy \citep{sympy},
          valgrind \citep{valgrind},
          VODE \citep{vode},
          yt \citep{yt}}

>>>>

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We study the coupling of hydrodynamics and reactions in simulations of
the double detonation model for Type Ia supernovae.  When assessing the
convergence of simulations, the focus is usually on spatial resolution, 
however, the method of coupling the physics together as well as the
tolerances used in integrating a reaction network also play an important role.
In this paper, we explore how the choices made in both coupling and integrating the reaction
portion of a simulation influences the accuracy, efficiency, and the nucleosynthesis
of simulations of the double detonation model of Type Ia supernovae.
We find that the stronger coupling between hydrodynamics and reactions,
afforded by the simplified spectral deferred corrections method introduced
previousl,y provides a more accurate and efficient result than standard
operator splitting.  We also find no need to limit reactions rates or reduce the
simulation timestep to the reaction timescale.  The entire simulation
methodology used here is freely available as part of the \castro\ simulation
code.
\end{abstract}

@@

There are many different progenitor models for Type Ia supernova (SN Ia) that
are currently being explored to explain the observed diversity of SN Ia explosions \cite{taubenberger:2017}.  The double detonation model
\citep{fink:2007} involves a sub-Chandrasekhar mass white dwarf with a
thin accreted He layer.  A detonation in the He layer propagates
around the star while sending a compression wave converging toward the
center.  This compression wave ignites a carbon detonation that then
propagates outward from the center of the white dwarf.  Many groups
have modeled this scenario, varying the initial conditions,
resolution \citep{rivas:2022}, and reaction networks to understand the robustness of this
model as well as its connection to observations.

\citep{townsley:2019} \citep{gronow:2020}, \citep{gronow:2021}, \citep{glasner:2018}, \citep{roy:2022}, \citep{shenbildsten:2014}, \cite{polin:2019}, \citep{kushnirkatz:2020}

These events are modeled using the Euler equations with reactive and
gravitational sources:
\begin{eqnarray}
\ddt{\rho} + \nabla \cdot (\rho \Ub) &=& 0 \\
\ddt{(\rho \Ub)} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &=& \rho \gb \\
\ddt{(\rho E)} + \nabla \cdot (\rho \Ub E + \Ub p) &=& \rho \Ub \cdot \gb + \rho \dot{S} \\
\ddt{(\rho X_k)} + \nabla \cdot (\rho \Ub X_k) &=& \rho \omegadot_k
\end{eqnarray}
where $\rho$ is the mass density, $\Ub$ is the velocity, $E$ is the specific
total energy, $p$ is the pressure, and $X_k$ are the mass fractions
of the nuclei that react.  The sources are the gravitational acceleration,
$\gb$, which is obtained by solving the the Poisson equation
\begin{equation}
\nabla^2 \Phi = 4 \pi G \rho
\end{equation}
(where $G$ is Newton's constant) for the gravitational potential, $\Phi$, and defining $\gb = -\nabla \Phi$.
The reaction network provides the species creation rate, $\omegadot$ and the energy generation rate, $\dot{S}$.
Finally, the system is closed via the equation of state:
\begin{equation}
p = p(\rho, e, X_k)
\end{equation}
where $e$ is the specific internal energy, obtained as $e = E - |\Ub|^2/2$.

Simulation codes use different methods to advance this system of equations.  For hydrodynamics,
explicit-in-time integration is usually used, with the timestep, $\Delta t$, restricted by the Courant limit.  In two-dimensions, this is:
\begin{equation}
\Delta t = C \min_{i,j} \left \{ \frac{\Delta x}{|u_{i,j}| + c_{i,j}}, \frac{\Delta y}{|v_{i,j}| + c_{i,j}}  \right \}
\end{equation}
where $\Delta x$ and $\Delta y$ are the grid width in the $x$- and $y$-directions, $u$ and $v$
are the $x$- and $y$-components of the velocity and $c$ is the sound
speed, and the minimum is taken over all zones on the grid.  Here, the
Courant number, $C$ is required to be $C < 1$ for stability.  We note that
this is the form of the stability criterion for the corner transport upwind
formulation \citep{ppmunsplit} of hydrodynamics we use.  Dimensionally-split
or method-of-lines formulations will have a slighly more restrictive criteria.
Reactions, on the other hand, tend to be stiff, and are usually evolved
using implicit-in-time integration, with many small steps taken to
make up the hydrodynamics timestep $\Delta t$.

@@

When using operator-splitting, the density remains constant, since there
is no reaction source in the mass continuity equation.  The mass
fractions, $X_k$, and energy evolve according to:
\begin{eqnarray}
\frac{dX_k}{dt} &= \omegadot_k(\rho, T, X_k) \label{eq:stranga} \\
\frac{de}{dt} &= \dot{S}(\rho, T, X_k) \\
T &= T(\rho, e, X_k) \label{eq:strangc}
\end{eqnarray}
In the simplest approximation of operator splitting, the
evolution of the mass fractions is solved alone, without integrating
the temperature or energy.  This is the form that was used, e.g., in FLASH
\citep{flash}.  A more accurate operator splitting also includes the
energy evolution (perhaps in terms of temperature), solving the system Eqs.~\ref{eq:stranga}--\ref{eq:strangc}.  This will require
calling the equation of state each time we evaluate the righthand side
of the ODE system, potentially increasing the computational expense.  However, as we showed in
\citet{strang_rnaas}, this is needed to get second-order convergence.

Detonations can be difficult to model numerically.  One common
approach is to use a simultion timestep based on the reaction
timescale, as discussed in \citet{prometheus}, instead of the
hydrodynamics timescale.  For the double detonation scenario, the
ignition of the second detonation can be very sensitive to how the
simulation is run \MarginPar{refs}.  Artifically limiting the
rates \MarginPar{ref}

In \citet{castro_simple_sdc}, we introduced a method based on the
ideas of spectral deferred corrections for coupling reactions and
hydrodyamics (that we called ``simplified-SDC'').  Here, the overall
time-integration is done iteratively, with the hydrodynamics seeing an
explicit reactive source, and the reaction update evolving an ODE
system that includes a piecewise-constant-in-time advective source,
$\Adv{\Uc}^{n+1/2}$,
\begin{equation}
\label{eq:simplesdc}
\ddt{\Uc} = \Rb(\Uc) + \Adv{\Uc}^{n+1/2}
\end{equation}
Here, $\Uc$ is the conserved state, $\Uc = (\rho, \rho \Ub, \rho E,
\rho X_k)^\intercal$, $\Rb(\Uc)$ are the reactive sources, and
$\Adv{\Uc}^{n+1/2}$ is an approximation to the advective update over
the timestep, as computed by the corner transport upwind formulation
of the piecewise parabolic method
\citep{ppmunsplit,millercolella:2002,ppm}. Crucially, this advective
term includes an explicit reaction source term and the update,
Eq.~\ref{eq:simplesdc} means that the reactions will ``see'' what
advection is doing as the reaction network is integrated.

In addition to the time-integration strategy, integrating the reaction
ODE system requires specifying tolerances, usually both an absolute, $\epsilon_\mathrm{abs}$, and relative tolerance, $\epsilon_\mathrm{rel}$, that are combined together into a weight of the form:
\begin{equation}
w_i = \epsilon_{\mathrm{abs},i} + \epsilon_{\mathrm{rel},i} |y_i|
\end{equation}
where the $y_i$ is one of the variables being integrated by the ODE
integrator and $i$ is the index, indicating that different tolerances
can be used for each variable.  Traditionally one set of tolerances is
used for the mass fractions and another for the energy (if it is
integrated).  Unfortunately, the tolerances used by simulations are
not normally reported in papers, and in many cases, may not be easily
controlled by a code user at runtime (and therefore, users may not be
aware of what they are using).  Some examples can be found in publicly
available networks.  For example, the freely available version of the
{\tt aprox13} network uses $\epsilon_\mathrm{rel} = 10^{-5}$ and a
floor on the mass fractions corresponding to $\epsilon_\mathrm{abs} =
10^{-6}$.  Since the simplified-SDC update works in terms of the
conserved variables, $\Uc$, we apply a density weighting to the absolute tolerances when comparing to the values used in Strang integration.

@@

\begin{figure}[t]
\centering
\plotone{subch_Temp_res_compare}
\caption{\label{fig:res_panel} Temperature evolution for SDC simulations with CFL=0.2 at 40, 20, and 10 km.  For the earliest time
  in each resolution, the AMR grid structure is shown.}
\end{figure}

@@

\software{\amrex~\citep{amrex_joss},
          \castro~\citep{castro,castro_joss},
          GCC (\url{https://gcc.gnu.org/}),
          helmeos \citep{timmes_swesty:2000},
          linux (\url{https://www.kernel.org/}),
          matplotlib (\citealt{Hunter:2007}, \url{http://matplotlib.org/}),
          NetworkX \citep{networkx},
          NumPy \citep{numpy,numpy2},
          pynucastro \citep{pynucastro,pynucastro2},
          python (\url{https://www.python.org/}),
          SymPy \citep{sympy},
          valgrind \citep{valgrind},
          VODE \citep{vode},
          yt \citep{yt}}
